---
bibliography: TestBiblio.bib
csl: biological-conservation.csl
output:
  word_document:
    fig_caption: yes
    reference_docx: DefaultMSstyle.docx
  pdf_document: default
  html_document: default
---

```{r, echo=FALSE, results='hide'}
library(pander, quietly=T)
library(knitr, quietly=T)

pasteLabel <- function(preText, inObj, objName, insLink=FALSE){
  objNum <- inObj[objName]
  useText <- paste(preText, objNum, sep=" ")
  if (insLink){
    useText <- paste("[", useText, "](#", objName, ")", sep="")
}
useText
}
 
# this increments the counter, and gives a name to the number so we can reference it later
incCount <- function(inObj, useName){
  nObj <- length(inObj)
  useNum <- max(inObj) + 1
  inObj <- c(inObj, useNum)
  names(inObj)[nObj+1] <- useName
  inObj
}
figCount <- c("_"=0)
tableCount <- c("_"=0)
 
tableCat <- function(inFrame){
    outText <- paste(names(inFrame), collapse=" | ")
    outText <- c(outText, paste(rep("---", ncol(inFrame)), collapse=" | "))
    invisible(apply(inFrame, 1, function(inRow){
    outText <<- c(outText, paste(inRow, collapse=" | "))
    }))
  return(outText)
} 
```

# Limited evidence for cumulative effects of small wind turbines on bat activity on a landscape scale
*Jeroen Minderman, Mairi Gillis, Helen Daly & Kirsty Park*

# 1. Introduction

In this paper we aim to quantify and analyse the potential cumulative effect of SWTs on bat activity on a wider landscape scale (up to 500m from installed turbines). Specifically, using data collected at **X sites** throughout the UK we test the following predictions:
(1) Bat activity is systematically lower in closer proximity (e.g. 0-100m from SWTs compared to 100-200m, 200-300m, etc) of operating SWTs, controlling for the effects of habitat and environmental conditions.
(2) The effect of SWT proximity on bat activity as tested in Prediction 1 is stronger in sites with multiple (2-4) SWTs installed compared to single SWT sites. Support for this prediction would indicate evidence of cumulative effects of SWTs.

# 2. Methods

General methods text.

## Bat data

Transects; length, orientation. Time of data collection. Details of detectors. 

## Habitat data

Briefly, quantification of all variables. Refer to appendix for details of "initial" habitat data model selection. 

## Statistics

You can also embed plots, for example:

```{r, echo=FALSE, fig.cap='This is a test figure.'}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## References {-}