---
bibliography: TestBiblio.bib
csl: biological-conservation.csl
output:
  word_document:
    fig_caption: yes
  html_document: default
---

# 1. Introduction

This is the introduction of the text. In fact, this is a random test of some stuff to do with a test bibliography what I just made [@collins_differences_2009]. This is a further test with two references [@martin_understanding_2011; @hooten_guide_2014]. As Herborn et al. [-@herborn_personality_2014] pointed out, tits are pretty cool.
This is a great big long block of text with a hard return after *this*.
Some further text, after which I will attempt a page break.



# 2. Methods
General methods text close to heaader.

## 2.1. Methods subsection 1

Here is some more methods text with a code block.

```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE, fig.cap='This is a test figure.'}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## References {-}