---
bibliography: TestBiblio.bib
csl: biological-conservation.csl
output:
  word_document:
    fig_caption: yes
    reference_docx: DefaultMSstyle.docx
  pdf_document: default
  html_document: default
---

```{r, echo=FALSE, results='hide'}
library(pander)
library(knitr)

pasteLabel <- function(preText, inObj, objName, insLink=FALSE){
  objNum <- inObj[objName]
  useText <- paste(preText, objNum, sep=" ")
  if (insLink){
    useText <- paste("[", useText, "](#", objName, ")", sep="")
}
useText
}
 
# this increments the counter, and gives a name to the number so we can reference it later
incCount <- function(inObj, useName){
  nObj <- length(inObj)
  useNum <- max(inObj) + 1
  inObj <- c(inObj, useNum)
  names(inObj)[nObj+1] <- useName
  inObj
}
figCount <- c("_"=0)
tableCount <- c("_"=0)
 
tableCat <- function(inFrame){
    outText <- paste(names(inFrame), collapse=" | ")
    outText <- c(outText, paste(rep("---", ncol(inFrame)), collapse=" | "))
    invisible(apply(inFrame, 1, function(inRow){
    outText <<- c(outText, paste(inRow, collapse=" | "))
    }))
  return(outText)
} 
```

# 1. Introduction

This is the introduction of the text. In fact, this is a random test of some stuff to do with a test bibliography what I just made [@collins_differences_2009]. This is a further test with two references [@martin_understanding_2011; @hooten_guide_2014]. As Herborn et al. [-@herborn_personality_2014] pointed out, tits are pretty cool.
This is a great big long block of text with a hard return after **this**.
Some further text, after which I will attempt a page break. Here is a further part-sentence in which I will test the insertion of a different type of reference [@doi=10.1006/gcen.2001.7749]. This is the next sentence.



# 2. Methods
General methods text close to heaader.

## 2.1. Methods subsection 1

Here is some more methods text with a code block.

```{r}
summary(cars)
```

Here is the same table but run through tableCat():

```{r, results='asis', echo=FALSE}
cat(tableCat(head(cars,5)), sep='\n')
```

```{r, results='hide', echo=FALSE}
mytab <- head(cars,5)
tableCount <- incCount(tableCount, "mytab")
```
And here is the same table (`r I(pasteLabel('Table', tableCount, 'mytab'))`) again, but now with a caption and label:
```{r, results='asis', echo=FALSE, tab.}
pandoc.table(mytab, caption=paste(I(pasteLabel('Table', tableCount, 'mytab')), '. This is the actual table caption',sep=''))
```

Here is a further attempt with the same table but using kable():
```{r echo=F}
kable(mytab, caption='test')
```


### 2.1.2. Header level 3

You can also embed plots, for example:

```{r, echo=FALSE, caption='This is a test figure.'}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## References {-}